% Set up the document
\documentclass[a4paper, 11pt,oldfontcommands]{memoir}

\usepackage[pdfpagemode={UseOutlines},bookmarks=true,bookmarksopen=true,
   bookmarksopenlevel=0,bookmarksnumbered=true,hypertexnames=false,
   colorlinks,linkcolor={black},citecolor={black},urlcolor={black},
   pdfstartview={FitV},unicode,breaklinks=true]{hyperref}

% Include any extra LaTeX packages required 
\usepackage[square, numbers, super, comma, sort&compress]{natbib} 
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{bussproofs}
\DeclareGraphicsExtensions{.eps}
\usepackage{tabularx}
\usepackage{array}
\let\footruleskip\undefined
\usepackage{fancyhdr}
\usepackage{float}
\usepackage{tikz}
\usepackage{fix-cm}
\usepackage{circuitikz}
\usetikzlibrary{arrows, automata, shadows}
\usepackage{todonotes}
\usepackage{paralist}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{polytable}
\RequirePackage{calc}
\usepackage{IEEEtrantools}
\usepackage{txfonts}
\usepackage[section]{placeins}
\usepackage[toc,style=list]{glossaries}
\usepackage{tcolorbox}

\DisemulatePackage{setspace}
\usepackage{setspace}
\onehalfspacing
\setlength{\parindent}{0pt}
\setlength{\parskip}{2.0ex plus0.5ex minus0.2ex}
\setstretch{1.35}  

\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\singlespacing}

\usepackage{vmargin}
\setmarginsrb           { 1.5in}  % left margin
                        { 0.6in}  % top margin
                        { 1.0in}  % right margin
                        { 0.8in}  % bottom margin
                        {  20pt}  % head height
                        {0.25in}  % head sep
                        {   9pt}  % foot height
                        { 0.3in}  % foot sep

\newenvironment{changemargin}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setlength{\parsep}{\parskip}%
}%
\item[]}{\end{list}}

\chapterstyle{bianchi}

%shorthands for style
\newcommand{\ttf}[1] {\texttt{#1}}
\newcommand{\tif}[1] {\textit{#1}}
\newcommand{\tbf}[1] {\textbf{#1}}

\tcbuselibrary{listings}
\tcbuselibrary{skins}
\newcounter{texexp}
\newcounter{mydef}
\newcounter{myexp}

\tcbset{
%
% Code Snippet
%
texexp/.style={%colback=red!50!yellow!5!white,
colframe=black!85,
colback=white,%gray!1,
enhanced,
colbacktitle=gray!33,
coltitle=black,
%listing options={language=haskell,columns=fullflexible,keywordstyle=\color{red},basicstyle=\small},
%fonttitle=\small\sffamily\bfseries, 
%fontupper=\small, fontlower=\small, 
fonttitle=\small,
arc=4mm,
outer arc=2.5mm,
%toptitle=1mm,
%bottomtitle=1mm,
left=0mm,right=0mm,top=0mm,bottom=0mm,
lefttitle=5mm,
leftrule=0mm,%0.75mm,
titlerule=0.25mm,
toprule=0.5mm,%0mm,
bottomrule=0.5mm, %0.25mm,
rightrule=0mm},%0.25mm},
example/.code 2 args={\refstepcounter{texexp}\label{#2}%
\pgfkeysalso{texexp,title={\textbf{Code Snippet \thetexexp:} #1}}},
%
% Definition
%
mydef/.style={%hcolback=red!50!yellow!5!white},
colframe=black!85,
colback=white,%gray!1,
enhanced,
colbacktitle=gray!33,
coltitle=black,
fonttitle=\small,
toptitle=0.25mm,
bottomtitle=0.25mm,
leftrule=0.25mm,%0.75mm,
titlerule=0.25mm,
toprule=0.5mm,%0mm,
bottomrule=0.5mm, %0.25mm,
rightrule=0.25mm},%0.25mm},
definition/.code 2 args={\refstepcounter{mydef}\label{#2}%
\pgfkeysalso{mydef,title={\textbf{Definition \themydef:} #1}}},
%
% Expansion
%
myexp/.style={%hcolback=red!50!yellow!5!white},
colframe=black!85,
colback=white,%gray!1,
fonttitle=\small,
%left=0mm,right=0mm,top=0mm,bottom=0mm,
leftrule=0.75mm,toprule=0mm,bottomrule=0mm,rightrule=0mm},
expansion/.code 2 args={\refstepcounter{myexp}\label{#2}%
\pgfkeysalso{myexp}},
}
\newenvironment{texexp}[1]{\tcblisting{texexp,#1}}{\endtcblisting}
%\newenvironment{expansion}[1]{\tcblisting{expansion,#1}}{\endtcblisting}
\newenvironment{texexptitled}[3][]{\tcblisting{example={#2}{#3},#1}}{\endtcblisting}
\newenvironment{definitiontitled}[3][]{\tcblisting{definition={#2}{#3},#1}}{\endtcblisting}
\newenvironment{expansiontitled}[3][]{\tcblisting{expansion={#2}{#3},#1}}{\endtcblisting}
\newenvironment{expansionno}[1]{\tcblisting{expansion,#1}}{\endtcblisting}
%glossary
\input{./glossary.tex}

\hypersetup{urlcolor=blue, colorlinks=true}  % Colours hyperlinks in blue, but this can be distracting if there are many links.

%So that todonotes do not overflow in the margins
\setlength{\marginparwidth}{2cm}

\makeheadrule{headings}{\textwidth}{\normalrulethickness}		% Add a rule on each normal page (no chapter)
\makefootrule{headings}{\textwidth}{\normalrulethickness}{0ex}
\makeevenfoot{headings}{\thepage}{}{}						% Add page numbers in the footers (outer edges)
\makeoddfoot{headings}{}{}{\thepage}
\makeevenhead{headings}{\leftmark}{}{}						% Add chapter and section marks (outer edges)
\makeoddhead{headings}{}{}{\rightmark}

\usepackage[sc]{mathpazo} % use mathpazo for math fonts
\usepackage{tgpagella} % but use tgpagella as main font
\usepackage[T1]{fontenc}

\normalfont

\newlength\drop
%% ----------------------------------------------------------------
\begin{document}

\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother

%\title{MEANDERINGS}
\begin{titlingpage}
  \begin{center}
    \textsc{Master's thesis} \\
    \vspace{0.5cm}
    \rule{\textwidth}{1pt} \\
    \vspace{0.5cm}
    \LARGE \textsc{Timed Types for Synchronous Hardware} \\
    \vspace{0.5cm}
    \large \textsc{Expressing Temporal Behaviour through the Type System of a\\Functional Language} \\
    \vspace{0.5cm} 
    \rule{\textwidth}{1pt} \\
    \vspace{7cm}
  \end{center} 
  \emph{Author:}\\ 
  Gerald Otter \\
  \newline
  \emph{Committee:}\\ 
  Dr. Ir. Jan Kuper\\
  Ir. Christiaan Baaij\\
  Dr. Ir. Philip H\"olzenspies\\
  Ir. Bert Molenkamp\\

  \begin{center}
    \vfill
    Computer Architecture for Embedded System (CAES)\\
    Faculty of EEMCS\\
    University of Twente\\
    \vspace{1cm}
    \today
  \end{center} 
%\drop = 0.1\textheight
%%\vspace*{\baselineskip}
%\vfill
%\hbox{%
%\hspace*{0.05\textwidth}%
%\rule{1pt}{\dimexpr\textheight-28pt\relax}%
%\hspace*{0.05\textwidth}%
%\parbox[b]{0.80\textwidth}{
%\vbox{%
%\vspace{\drop}
%{\noindent\HUGE\bfseries Timed Types}\\[2\baselineskip]
%{\Large\itshape Expressing Timing Constraints through the Type System of a Functional Language}\\[.37\baselineskip]
%{\Large Gerald Otter}\par
%\vspace{0.5\textheight}
%{\noindent Committee:}\\
%{\noindent Dr. Ir. Jan Kuper}\\
%{\noindent Ir. Bert Molenkamp}\\
%{\noindent \begin{center} \today \end{center}}
%}% end of vbox
%}% end of parbox
%}% end of hbox
%\vfill
%\null
\end{titlingpage}

\newcommand*{\pretoctitle}[1]{{\clearpage\centering
   \vspace*{-\topfiddle}#1\par}}

\frontmatter	  % Begin Roman style (i, ii, iii, iv...) page numbering 
% Set up the Title Page 
%\title{}
%\author{G.J. Otter}

%\begin{center}
%\begin{tcolorbox}[enhanced,toprule=0mm,bottomrule=0mm,arc=4mm,outer arc=2.5mm,lowerbox=ignored,colback=white]
%{ \huge $\textsc{C\lambdaaSH_\tau}$ }
%\end{tcolorbox}
%\end{center}

\cleardoublepage
\chapter{Abstract}
As shown through the development of the functional hardware description language \gls{clash}, functional languages are well suited to describe synchronous hardware.
Functional languages allow creation of extensive type systems.
The type systems of functional languages can be used to verify vastly different properties of expressions.
Even though \gls{clash} is not the first functional hardware description language, none of the existing functional hardware description languages conscript the aid of the type system to verify properties specific to hardware designs.
Even though hardware description languages allow verification of certain properties during simulation, static analysis through the type system is only performed similarly to general purpose languages.

In this thesis, a type system is developed which allows verification of temporal properties of synchronous hardware specifications.
Aside from providing a formal definition of the type system through typing rules and grammar, an implementation is discussed which shows the practical feasibility of expressing temporal behaviour through the type system.
We show that descriptions of temporal behaviour through the type system leads to concise specifications, where the computational aspect of hardware definitions is separated from the temporal aspect.
We provide an analysis of the constraint-based approach used to create the type system, and determine what the pitfalls are when trying to include time in the type system. 

From the provided analysis we conclude that expressing time as part of the type system certainly has its advantages.
Specifications including time-dependent behaviour can be made more concise, though it does come at a cost of some impredictability when creating compositions.
However, the constraint-based type system discussed is not ideal for reasoning about the properties of time-dependent specifications, and is difficult to extend.
As a result, more research is definitely needed in order to more completely asses the possibility of expressing time through the type system.

\cleardoublepage
\chapter{Acknowledgements}
This thesis would not have been possible without the help of many.
Foremost I would like to thank my first supervisor Jan Kuper.
Aside from ample constructive criticism with regards to the content of this thesis, he has been paramount in guiding me through the forest of newly aquired knowledge.
Not only that, he has also generously given me personal advice when I needed it, even it was not always what I wanted to hear.
 
I would also like to thank Christiaan Baaij.
Christiaan has diligently read most of the texts I have produced over the past six months, which were admittedly difficult to interpret most of the time.
Without his help, even when disregarding \gls{clash}'s existence, this thesis would not be what it is today.
His experience with code generation and hardware synthesis has provided me with the necessary criticism to relate my ideas to practice.
Without his capacity to point out errors in logic and reasoning, the text before you would certainly be more difficult to decipher.

I feel lucky to have had Philip H\"olzenspies join in providing me with criticism during the final stretch of writing this thesis.
Without his help and strong criticism, the type system might not have existed at all, let alone in its current incarnation.
His advice showed me the value of my work, regardless of its results.

Over the past ten years, many fellow students have influenced the way I see the world and react to it.
In particular I would like to thank Mark de Ruiter, for the interesting conversations we had, even though we perhaps should have been working on our theses instead.
In that same vein, I would like to thank the people of the CAES group, for providing a fun and relaxed working environment, although I could have used a less relaxed environment at times.
I also thank my good friends from across the globe for their continuing support. 

Finally, I thank my immediate family. 
Thank you for supporting me, regardless of the goals I set for myself.
Thank you, my dearest mother and brother, and despite everything, my loving father.

\cleardoublepage
%% ----------------------------------------------------------------

% Define the page headers using the FancyHdr package and set up for one-sided printing
\fancyhead{}  % Clears all page headers and footers
\rhead{\thepage}  % Sets the right side header to show the page number
\lhead{}  % Clears the left side page header

%\pagestyle{fancy}  % Finally, use the "fancy" page style to implement the FancyHdr headers
%% ----------------------------------------------------------------
\tableofcontents  % Write out the Table of Contents
\cleardoublepage
%% ----------------------------------------------------------------
%\todototoc
%\listoftodos
%\cleardoublepage
%% ----------------------------------------------------------------
%\listoffigures  % Write out the List of Figures
%\cleardoublepage
%% ----------------------------------------------------------------
%\listofsources
%\cleardoublepage
%% ----------------------------------------------------------------

%\lhead{\emph{Acronyms}}  % Set the left side page header to "Abbreviations"

%\printglossaries

%% ---------------------------------------------------------------- 
\mainmatter	  % Begin normal, numeric (1,2,3...) page numbering

%\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
%\renewcommand{\sectionmark}[1]{\markright{#1}{}}
%
%\fancyhead[LE, RO]{\slshape \thepage} 
%\fancyhead[LO]{\thesection.  \slshape  \rightmark}
%\fancyhead[RE]{Chapter \thechapter. \slshape \leftmark}

\input{./chapters/introduction.lhs}
\input{./chapters/background.lhs}
\input{./chapters/syntax.lhs}
\input{./chapters/formalization}
\input{./chapters/implementation.lhs}
\input{./chapters/conclusion}
\input{./chapters/discussion.lhs}

\clearpage 
%% ----------------------------------------------------------------

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

\appendix % Cue to tell LaTeX that the following 'chapters' are Appendices

\lhead[\slshape \thepage]{Appendix \thechapter. \slshape \leftmark}
%\rhead[Appendix \thechapter. \slshape \leftmark]{\slshape \thepage}
\input{./chapters/existinglang.lhs}
%\input{./appendices/experimentSetup}

\addtocontents{toc}{\vspace{2em}}  % Add a gap in the Contents, for aesthetics
\backmatter

%% ----------------------------------------------------------------
\label{Bibliography}
\lhead[\slshape \thepage]{\slshape Bibliography}  % Change the left  side page header to "Bibliography"
\rhead[\slshape Bibliography]{\slshape \thepage}  % Change the right side page header to "Bibliography"
\bibliographystyle{plainnat}

\bibliography{chronologic}
\clearpage

\end{document}  % The End
%% ----------------------------------------------------------------

